# `ggplot2` Exercises {#ggplot-exercises .unnumbered}

## `ggplot2` Extended Example:  Walasek & Stewart (2015) Exp. 1a & 1b {#ggplot-walasek-example}

```{r, message=FALSE}
library("tidyverse")
ws1a <- read_csv("data/ws2015_exp1a.csv")
ws1b <- read_csv("data/ws2015_exp1b.csv")
ws1a <- ws1a %>%
  mutate(subno = factor(paste0("e1a_", subno)))
ws1b <- ws1b %>%
  mutate(subno = factor(paste0("e1b_", subno)))
ws1 <- bind_rows(ws1a, ws1b)
ws1 <- ws1 %>% 
  mutate(
    response = factor(response, levels = c("reject", "accept")),
    condition = factor(
      condition, 
      levels = c(40.2, 20.2, 40.4, 20.4), 
      labels = c("-$20/+$40", "-$20/+$20", "-$40/+$40", "-$40/+$20")
    )
  )
glimpse(ws1)
```

only the symmetric lotteries that are shared across conditions separately for each participant and condition.


```{r}
part_sum <- ws1 %>%
  filter(loss == gain, loss %in% c(12, 16, 20)) %>% 
  group_by(condition, subno) %>% 
  summarise(mean_acc = mean(resp)) %>% 
  ungroup()
part_sum
```

```{r}
ggplot(part_sum, aes(x = condition, y = mean_acc)) +
  geom_point()
```

```{r}
ggplot(part_sum, aes(x = condition, y = mean_acc)) +
  geom_jitter(width = 0.2, height = 0.1, alpha = 0.25)
```

```{r}
ggplot(part_sum, aes(x = condition, y = mean_acc)) +
  geom_jitter(width = 0.2, height = 0.1, alpha = 0.25)
```



## Quiz 

Note: The pull-down menu for selecting an answer turns green after selecting the correct answer. 

::: {.exercise}
Which variable is typically plotted on the y-axis? 

1. The independent variable
2. The dependent variable
3. The control variable

Answer: `r webex::mcq(c("1", answer = "2", "3"))`
:::


::: {.exercise}
Consider you have data from an experiment with dependent variable `score` and  independent variable `condition`. What would be the `aes()` call to plot this data in the conventional way?

1. `aes(condition = x, score = y)`
2. `aes(condition = y, score = x)`
3. `aes(x = condition, y = score)`
4. `aes(y = condition, x = score)`

Answer: `r webex::mcq(c("1", "2", answer = "3", "4"))`
:::


::: {.exercise}
What is the appropriate way to handle outliers that are identified by a box-plot?

1. Outliers should be removed from the analysis as they can influence the results to a large degree.
2. Any real data point that is collected needs to be included in the analysis. Removing an outlier can be seen as an instance of data fabrication. 
3. We should repeat the analysis with and without outlier, to see if the results are robust to the presence or absence of the outlier.

Answer: `r webex::mcq(c("1", "2", answer = "3"))`
:::

