---
title: "Introduction to Statistics for Experimental Psychology with `R`"
author: "Henrik Singmann"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "Statistics for Experimental Psychology with R using the afex package."
---

```{r setup, include=FALSE}
library("tidyverse")
library("afex")
library("ggbeeswarm")
theme_set(theme_bw(base_size = 15) + 
            theme(legend.position="bottom", 
                  panel.grid.major.x = element_blank()))
```

# Overview

At this point in time, the book is being written so only few chapters are already available.

-   Chapter \@ref(role-of-statistics-in-the-research-process) provides a general introduction.
-   Chapter \@ref(standard1) introduces the basic statistical approach.

## Acknowledgments

This project would not exist without the help and feedback provided by others: David Kellen, Lukasz Walasek, Stuart Rosen

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
