---
title: "Introduction to Statistics for Experimental Psychology with `R`"
author: "Henrik Singmann"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "Statistics for Experimental Psychology with R using the afex package."
---

```{r setup, include=FALSE}
library("tidyverse")
library("afex")
library("ggbeeswarm")
theme_set(theme_bw(base_size = 15) + 
            theme(legend.position="bottom", 
                  panel.grid.major.x = element_blank()))
```

# Overview

At this point in time, the book is being written so only few chapters are already available.

-   Chapter \@ref(role-of-statistics-in-the-research-process) provides a general introduction.
-   Chapter \@ref(standard1) introduces the basic statistical approach.

## Acknowledgments

This project would not exist without the help and feedback provided by others: David Kellen, Lukasz Walasek, Stuart Rosen, Anna Krason

## License and Attribution

This book is licensed under the [Attribution-NonCommercial 4.0 International (CC BY-NC 4.0)](https://creativecommons.org/licenses/by-nc/4.0/) license. This license allows you to share and adapt the work as long as you give appropriate attribution and do not use the materials for commercial purposes.

Parts of this book uses other materials released under a compatible CC license. Where it does so, the source is clearly indicated. Please ensure to attribute the original source in case you re-use such materials. 


```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
